# Digits recognition on PNG images. R code

1. Added utility functions
2. Introspection on dim=2. Extra layers on tensor
3. Converting code from Python to R

## Objectives
1. Read PNG images instead of MNIST standard dataset.
2. Save the model
3. Read an unseen digit as PNG file using the saved model
4. Predict the digit and calculate accuracy

```{r, rsuite_setup}
# Detect proper script_path (you cannot use args yet as they are build with tools in set_env.r)
script_path <- (function() {
  args <- commandArgs(trailingOnly = FALSE)
  script_path <- dirname(sub("--file=", "", args[grep("--file=", args)]))
  if (!length(script_path)) {
    return(".")
  }
  if (grepl("darwin", R.version$os)) {
    script_path <- gsub("~\\+~", " ", script_path) # on MacOS ~+~ in path denotes whitespace
  }
  return(normalizePath(script_path))
})()

# Setting .libPaths() to point to libs folder
source(file.path(script_path, "set_env.R"), chdir = T)

config <- load_config()
args <- args_parser()
```

```{r, prj_paths}
proj_root_path <- file.path(script_path, "..", "..")
conda_path <- file.path(proj_root_path, "conda")
data_path <- file.path(proj_root_path, "data")
```


```{r, echo=FALSE}
# Force using local Python environment
if (.Platform$OS.type == "unix") {
  reticulate::use_python(python = file.path(conda_path, "bin",
                                            "python3"), require = TRUE)
} else if (.Platform$OS.type == "windows") {
  reticulate::use_python(python = conda_path,
                         require = TRUE)
}
# Find the PyQt libraries under Library/plugins/platforms
qt_plugins <- file.path(conda_path,
                        "Library", "plugins", "platforms")
reticulate::py_run_string("
import os;
os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = r.qt_plugins")
```


## Load PyTorch libraries

```{r}
library(reticulate)
library(testthat)

torch       <- import("torch")
torchvision <- import("torchvision")
nn          <- import("torch.nn")
transforms  <- import("torchvision.transforms")
dsets       <- import("torchvision.datasets")
builtins    <- import_builtins()
np          <- import("numpy")
```


## Dataset iteration batch settings

```{r}
# batch size for datasets
batch_size_train = 100L    # this will be used during training
batch_size_test  = 1000L

# folders where the images are located
train_data_path = sprintf("%s/mnist_png_full/training/", data_path)
test_data_path  = sprintf("%s/mnist_png_full/testing/", data_path)
```


## Read datasets
This is not your typical MNIST digit dataset. The standard dataset is in IDX format where you cannot see the image directly from your browser. I prefer to use PNG images. That's why we have to specify the `train_data_path` and the `test_data_path`. The size of the PNG dataset is 19.3 MB which includes the 70,000 samples for training and testing.

First we will start by reading the images as raw as they come. Later, after getting the first accuracy numbers, we will try with normalization and see how much contributes in improving the accuracy.

```{r}
# read the datasets without normalization
train_dataset = torchvision$datasets$ImageFolder(root = train_data_path, 
    transform = torchvision$transforms$ToTensor()
)

test_dataset = torchvision$datasets$ImageFolder(root = test_data_path, 
    transform = torchvision$transforms$ToTensor()
)

# number of images for training and testing
loginfo("%s %s", py_len(train_dataset), py_len(test_dataset))
# another way of getting the number of elements in the training dataset
train_dataset$`__len__`()

loginfo("%s", class(train_dataset))       # <class 'torchvision.datasets.folder.ImageFolder'>

# in Python the object train_dataset[0] is a tuple
# In R, this will produce an error
expect_error(class(train_dataset[0]))      # <class 'tuple'>

# how does the train_dataset object presents itself
train_dataset
# Dataset ImageFolder
#     Number of datapoints: 60000
#     Root location: /home/msfz751/mnist_png_full/training/
```

## Read image and label for a data point
The Python object is a tuple with two elements but in R returns as a list of two elements: 
* [[1]] a tensor of size 3x28x28 rerpresenting the image, and
* [[2]] an integer, representing the label for the digit

```{r}
# retrieve a random data point from the dataset
random <- import("random")

# a random number from 0 through 59999
i_rand = random$randrange(0, py_len(train_dataset)-1)
i_rand

# the Python object is a tuple with two elements
# But in R returns as a list of two elements: 
#     [[1]] a tensor, and [[2]] an integer
dataset_object = py_get_item(train_dataset, i_rand) 
dataset_object
```

```{r}
# class and length of the dataset_object
class(dataset_object)
length(dataset_object)

# decomposing the dataset_object
image <- dataset_object[[1]]
label <- dataset_object[[2]]
```

The size of the image tensor is `torch.Size([3, 28, 28])` but it should really be `torch.Size([1, 28, 28])`.
```{r}
# this is just one of the 60,000 image-label objects
image$size()  # torch.Size([3, 28, 28])
```


```{r}
# the object label is not a tensor but rather an integer
# applying the size() function will throw an error.
builtins$isinstance(label, builtins$int)
expect_error(label$size())

# length of the integer as a string, same as number of digits
loginfo("Length of label: %d", builtins$len(builtins$str(label)))
```

### Introspection of the train_dataset

```{r}
loginfo("The object train_dataset is not directly accesible from R. 
Using the `py_get_item()` function we can reach any of its %d elements of the 
%s. Each of these objects has %d members.\n", py_len(train_dataset), builtins$type(py_get_item(train_dataset, 0L)), 
builtins$len(py_get_item(train_dataset, 0L))
)

loginfo("\tThe first member is of the %s, a tensor for the image.\n", 
            builtins$type(py_get_item(train_dataset, 0L)[[1]]))

loginfo("\tThe second member is of the %s, an integer for the label.",
            builtins$type(py_get_item(train_dataset, 0L)[[2]]))
```

Each of the images read from the PNG files is loaded into a tensor of size 3x28x28.

```{r}
# check size of first and last data point tensor, image part
py_get_item(train_dataset, 0L)[[1]]$size()      # first object
py_get_item(train_dataset, 59999L)[[1]]$size()  # last object
image$size()
# it is torch.Size([3, 28, 28]) but it should be 1, 28, 28
# we will take only one slice of dim=1
```

### Size of tensor on dim=2 of train_dataset
The tensors on dim=2 seem to be identical. We will confirm in the following chunk:


### Visualizing the images from the tensor

```{r}
# retrieve a random data point from the dataset
random <- import("random")

# a random number from 0 through 59999
i_rand = random$randrange(0, py_len(train_dataset)-1)
i_rand

# the Python object is a tuple with two elements
# But in R returns as a list of two elements: 
#     [[1]] a tensor, and [[2]] an integer
dataset_object = py_get_item(train_dataset, i_rand) 

# decomposing the dataset_object
image <- dataset_object[[1]]
label <- dataset_object[[2]]

image[0]$size()
```

```{r}
rotate <- function(x) t(apply(x, 2, rev))   # function to rotate the matrix

# convert to numpy array and reshape
img_np_rs = np$reshape(image[0]$numpy(), c(28L, 28L))
image(rotate(img_np_rs))
title(label)  # the label for the corresponding tensor
```

### Retrieve a second data point

```{r}
# second random data point
i_rand = random$randrange(0, py_len(train_dataset)-1)   # get a random data point

dataset_object = py_get_item(train_dataset, i_rand)    # read the tuple
image <- dataset_object[[1]]
label <- dataset_object[[2]]

# convert to numpy array and reshape
img_np_rs = np$reshape(image[0]$numpy(), c(28L, 28L))
image(rotate(img_np_rs))
title(label)  # the label for the corresponding tensor
```

## Reduce the number of layers for dim=1 in the image

```{r}
# this class to be used to get rid of two duplicate layers in the image
main <- py_run_string('
class PickLayerTransform:
    def __init__(self, layer):
        # self.img_ds = img
        self.layer = layer
        if self.layer < 0: raise RuntimeError("Layer index {} cannot be negative ".format(self.layer))
        
    def __call__(self, img):
        if (self.layer > len(img)-1): raise RuntimeError("Layer index {} incompatible with dimension size {}".format(self.layer, len(img)))
        return img[(self.layer-1):self.layer, :, :]
')

PickLayerTransform <- main$PickLayerTransform

# trying to resize tensor to [1, 28, 28]
train_dataset = torchvision$datasets$ImageFolder(root = train_data_path,
    transform = torchvision$transforms$Compose(c(
              transforms$ToTensor(),
              PickLayerTransform(1L)
    )))

test_dataset = torchvision$datasets$ImageFolder(root = test_data_path,
    transform = torchvision$transforms$Compose(c(
              transforms$ToTensor(),
              PickLayerTransform(1L)
    )))
```

```{r}
# check size of a data point tensor, image part
# train_dataset[0][0].size()       # first object
# train_dataset[59999][0].size()   # last object
first <- 0L
last <- 59999L
loginfo("New dataset image size:\n     %d of %d: %s.\n %d of %d: %s.\n", 
  first+1, last+1, py_get_item(train_dataset, first)[[1]]$size(),
  last+1,last+1, py_get_item(train_dataset, last)[[1]]$size()
)

# this image generated before eliminating 2 extra layers
loginfo("Old dataset image size:\n %s",
            image$size()
)
```

> It  was torch.Size([3, 28, 28]) but it should have been torch.Size([1, 28, 28]). 

### Show a random image

```{r}
i_rand = random$randrange(0, py_len(train_dataset)-1)   # get a random data point

dataset_object <- py_get_item(train_dataset, i_rand)    # read the list

# extract image and label from the list
image <- dataset_object[[1]]
label <- dataset_object[[2]]

# convert to numpy array and reshape
# we don't need to specify the layer index anymore
# we can use this form:
  # img_np_rs = img[:, :, :].numpy().reshape(28, 28) in Python
  # or this in R:
img_np_rs = np$reshape(image$numpy(), c(28L, 28L))
image(rotate(img_np_rs))
title(label)  # the label for the corresponding tensor
```

### Apply `DataLoader` and set the batch size

To prevent losing features by using a simple for loop to iterate over the data. In particular, we are missing out on:

    Batching the data
    Shuffling the data
    Load the data in parallel using multiprocessing workers.


```{r}
# load the dataset of images
train_loader = torch$utils$data$DataLoader(
        train_dataset,
        batch_size=batch_size_train,
        shuffle=TRUE
    )

# load the dataset of images
test_loader = torch$utils$data$DataLoader(
        test_dataset,
        batch_size=batch_size_test,
        shuffle=TRUE
    )

loginfo("%s %s", py_len(train_loader), py_len(test_loader))
```


```{r}
# Confirm that the dataset loaders are iterable objects
collections <- import("collections")

builtins$isinstance(train_loader, collections$Iterable)
builtins$isinstance(test_loader, collections$Iterable)
```

## Build the model

```{r}
# Build the model
# Same as linear regression! 
main <- py_run_string("
import torch.nn as nn

class LogisticRegressionModel(nn.Module):
    def __init__(self, input_dim, output_dim):
        super(LogisticRegressionModel, self).__init__()
        self.linear = nn.Linear(input_dim, output_dim)

    def forward(self, x):
        out = self.linear(x)
        return out
")

LogisticRegressionModel <- main$LogisticRegressionModel
```



```{r}
# feeding the model with 28x28 images
input_dim = 28L*28L

# classify digits 0-9 a total of 10 classes,
output_dim = 10L

# instantiate model
model = LogisticRegressionModel(input_dim, output_dim)
model
```

```{r}
### Instantiate Cross Entropy Loss class
# need Cross Entropy Loss to calculate loss before we backpropagation
criterion = nn$CrossEntropyLoss()  

# calculate parameters' gradients and update
learning_rate = 0.001

### Instantiate Optimizer class
optimizer = torch$optim$SGD(model$parameters(), lr=learning_rate)  
optimizer
```

```{r}
# Type of parameter object
loginfo("%s", model$parameters())
model_parameters <- builtins$list(model$parameters())

# Length of parameters
loginfo("%s", builtins$len(model_parameters))

# FC 1 Parameters 
loginfo("%s", builtins$list(model_parameters)[[1]]$size())

# FC 1 Bias Parameters
loginfo("%s", builtins$list(model_parameters)[[2]]$size())
```

We arbitrarily set 3000 iterations here which means the model would update 3000 times. 

```{r}
n_iters = 3000L
```

One epoch consists of 60,000 / 100 = 600 iterations. Because we would like to go through 3000 iterations, this implies we would have 3000 / 600 = 5 epochs as each epoch has 600 iterations. 

```{r}
num_epochs = n_iters / (py_len(train_dataset) / batch_size_train)
num_epochs = as.integer(num_epochs)
num_epochs
```

## Training the model
We have three loops: 
1. A loop for the epochs
2, A loop for the training dataset
3. A loop that evaluates accuracy on the test dataset after each epoch

```{r, r_training_model}
# train the model
iter <- 0L
for (epoch in 1:num_epochs) {
    iter_train_dataset <- builtins$enumerate(train_loader) # reset iterator
    for (train_obj in iterate(iter_train_dataset)) {
        # extract images, labels
        images <- train_obj[[2]][[1]]
        labels <- train_obj[[2]][[2]]
        # Load images as Variable
        images = images$view(-1L, 28L*28L)$requires_grad_()
        labels = labels
        # Clear gradients w.r.t. parameters
        optimizer$zero_grad()
        # Forward pass to get output/logits
        outputs = model(images)
        # Calculate Loss: softmax --> cross entropy loss
        loss = criterion(outputs, labels)
        # Getting gradients w.r.t. parameters
        loss$backward()
        # Updating parameters
        optimizer$step()
        iter = iter + 1
        if (iter %% 500 == 0) {
            # Calculate Accuracy for each epoch        
            correct <- 0
            total <- 0
            # Iterate through test dataset
            iter_test_dataset <- builtins$enumerate(test_loader) # reset iterator
            for (test_obj in iterate(iter_test_dataset)) {
                # Load images to a Torch Variable
                images <- test_obj[[2]][[1]]
                labels <- test_obj[[2]][[2]]
                images <- images$view(-1L, 28L*28L)$requires_grad_()
                # Forward pass only to get logits/output
                outputs = model(images)
                # Get predictions from the maximum value
                .predicted = torch$max(outputs$data, 1L)
                predicted <- .predicted[1L]
                # Total number of labels
                total = total + labels$size(0L)
                # Total correct predictions
                correct = correct + sum((predicted$numpy() == labels$numpy()))
            }
            accuracy = 100 * correct / total
            # Print Loss
            loginfo('Iteration: %5d. Loss: %f. Accuracy: %8.2f \n', 
                  iter, loss$item(), accuracy)
        }
    }
}
```

> This is a modified version of the original article.

Source:  https://www.deeplearningwizard.com/deep_learning/practical_pytorch/pytorch_logistic_regression/
